---
title: "Project1"
author: "Sophie Kikuchi, Ethan Buckner, and Olivia Conner"
date: "2023-09-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preliminary manipulation of the data
```{r}
library(tidyverse)
library(dplyr)
```

```{r}
house <- read.csv("house_total.csv")
```

## Data Cleaning
```{r}
house1 <- house %>%
  filter(party %in% c("DEMOCRAT", "REPUBLICAN"))

house1
```

```{r}
house1 <- subset(house1, select = -c(state_fips, state_cen, state_ic, office, stage, runoff, special, writein, mode, unofficial, version, fusion_ticket))

```

```{r}
house2 <-house1 %>%
  group_by(year, state, district) %>%
  summarize(
    democrat_votes = sum(ifelse(party == "DEMOCRAT", candidatevotes, 0)),
    republican_votes = sum(ifelse(party == "REPUBLICAN", candidatevotes, 0)),
    totalvotes = totalvotes,
    democrat_candidate = first(candidate[party == "DEMOCRAT"]),
    republican_candidate =first(candidate[party == "REPUBLICAN"])
  ) %>%
  ungroup()
```
```{r}
house_unique <- house2 %>%
  distinct()
```

```{r}
write.csv(house_unique, "house_clean.csv", row.names = FALSE)
```

## Creation of New Columns 
```{r}

# Finding margin of votes (positive = Democrat win, negative = Republican win)
house_unique <- house_unique %>%
  rowwise() %>%
  mutate(excess_dem_votes = democrat_votes - republican_votes)

# Creating column with total number of votes
house_unique <- house_unique %>%
  rowwise() %>%
  mutate(total_demrep_votes = democrat_votes + republican_votes)

# Creating a column of democrativ margin
house_unique <- house_unique %>%
  rowwise() %>%
  mutate(dem_margin = excess_dem_votes/total_demrep_votes)

# Eliminating rows where a candidate ran unopposed
house_filtered <- house_unique %>%
  filter(dem_margin != 1 & dem_margin != -1)

write.csv(house_filtered, "house_clean_new.csv", row.names = FALSE)

```

## Creating a "winner" Column
```{r}
house_unique <- house_unique %>%
  rowwise() %>%
  mutate(winner = ifelse(democrat_votes >= republican_votes, "Democrat", "Republican"))


# Print the result
print(house_clean)
```

## Adding a column to the main dataset showing the percentage of votes that the winning candidate had
```{r}

house_unique$winner_percentage <- ifelse(house_unique$winner == "Democrat", house_unique$democrat_votes / house_unique$totalvotes, house_unique$republican_votes / house_unique$totalvotes)

```

# Exploratory Analysis

## Filtering the total data set to only show from 2012 - 2020 data (fit in with our question)
```{r}
house_unique_recent <- house_unique %>%
  filter(year >= 2012, year <= 2020)
```

## Finding the number of repeat elected candidates by Party (democrat)
```{r}

# Filter rows where the winner is "Democrat"
democrat_winners <- house_unique_recent[house_unique_recent$winner == "Democrat", ]

# Calculate the count of wins for each Republican candidate
win_countd <- table(democrat_winners$democrat_candidate)

avg_percentage_d <- tapply(democrat_winners$winner_percentage, democrat_winners$democrat_candidate, mean)

# Create a data frame with candidate names and their win counts
result_data_d <- data.frame(
  candidate_name = names(win_countd),
  win_count = as.vector(win_countd),
  avg_percentage = avg_percentage_d
)

write.csv(result_data_d, "house_democrat_imcumbent.csv", row.names = FALSE)

```

## Finding the number of repeat elected candidates by Party (republican)
```{r}

# Filter rows where the winner is "Republican"
republican_winners <- house_unique_recent[house_unique_recent$winner == "Republican", ]

# Calculate the count of wins for each Republican candidate
win_countr <- table(republican_winners$republican_candidate)

avg_percentage <- tapply(republican_winners$dem_margin, republican_winners$republican_candidate, mean)

# Create a data frame with candidate names and their win counts
result_data_r <- data.frame(
  candidate_name = names(win_countr),
  win_count = as.vector(win_countr),
  avg_percentage = avg_percentage
)

write.csv(result_data_r, "house_republican_imcumbent.csv", row.names = FALSE)
```

## Graph of Average Percentage points that Democrats won by vs. # of repeated wins
```{r}

average_by_wins_d <- aggregate(avg_percentage ~ win_count, result_data_d, mean)

# Create a bar graph using ggplot2
ggplot(average_by_wins_d, aes(x = win_count, y = avg_percentage)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(
    title = "Average Percentage Points vs. Number of Repeat Wins for Democrat Party",
    x = "Number of Repeat Wins",
    y = "Average Percentage Points that Democrats won by"
  ) +
  theme_minimal()
```

## Graph of Average Percentage points that Democrats lost by vs. # of repeated wins
```{r}
average_by_wins_r <- aggregate(avg_percentage ~ win_count, result_data_r, mean)


# Create a bar graph using ggplot2
ggplot(average_by_wins_r, aes(x = win_count, y = avg_percentage)) +
  geom_bar(stat = "identity", fill = "red") +
  labs(
    title = "Average Percentage Points vs. Number of Repeat Wins for Republican Party",
    x = "Number of Repeat Wins",
    y = "Percentage points that Democrats lost by"
  ) +
  theme_minimal()
```


## Calculating leans by state and district for all years
```{r}
state_leans_by_district <- house_filtered %>%
  group_by(state, district) %>%
  summarize(avg_dem_margin = mean(dem_margin, na.rm = TRUE))

write.csv(state_leans_by_district, "house_leans_all.csv", row.names = FALSE)
```
## Calculating leans by state and district for years 2012-2018 
```{r}
state_leans_by_district_2012 <- house_filtered %>%
  filter(year >= 2012 & year <= 2020) %>%
  group_by(state, district) %>%
  summarize(avg_dem_margin = mean(dem_margin, na.rm = TRUE))

write.csv(state_leans_by_district_2012, "house_leans_recent.csv", row.names = FALSE)
```


# Extra Code

## Data set with just the 2022 Data
```{r}
house_2022 <- house_unique %>%
  filter(year == 2022)
```

